 

### 2023-07-18 markdown/methods.Rmd
until batch correction done

## Wetlab - Materials and Methods
For testing and development of the ProteoScanR pipeline the following materials and methods were performed.

### Celltypes
Macrophages from 12 healthy donors and 12 HIV patients with metabolic syndrome and 12 HIV patients without metabolic syndrome. Cell sorting was performed with FACSAria Fusion (BD Biosciences).

@@

### Labeling
Samples of test dataset were labeled with 12-plex TMTpro (Thermo Fisher) in anhydrous acetonitril (ACN) solution.

@@

### Hardware
Lenovo Laptop X360 with:
Intel(R) Core(TM) i5-6200U CPU @ 2.30GHz
8 GB RAM

@@

### Filter by q-value
The next step for quality control is the exclusion of samples with a high false discovery rate (FDR). When applying multiple statistical testing (e.g. t-Test) the obtained p-values can be considered as biased, because the probability to observe a significant will iteratively increase with each test performed. Corrections in statistics are an approach to compensate for the multiplicity of testing. There are many ways to do this compensation like the Bonferroni method or Benjamini-Hochberg`s FDR. In Mass spectrometry the common “way to go” is calculating a false discovery rate, by dividing false PSMs (=hit of the decoy database) through the total number of PSMs above the peptide-spectrum matching score. The peptides spectrum matching score is defined as $-10 * log(p)$. Whereas the p-value is defined that the hit is done by chance. The calculation of the score is highly dependent on the data acquisition method used. MaxQuant uses Andromeda, an integrated search engine. Proteome Discoverer from Thermo Fisher utilizes different engines such as Mascot or Minora. As published by J.Cox in 2011 Mascot and Andromeda showed similar performance when comparing FDR values as a function of coverage. However the observed performance can be lower when dealing with a decreased coverage \citep{Cox2011}.The threshold for accepting an FDR of an individual PSM is described as q-value.

### Peptide spectrum match (PSM) aggregation to peptides 
In data science, aggregation refers to a row-wise operation that merges data based on a particular column using a specific function. In the context of processing from PSMs to unique peptides, the desired column is the PSM sequence. To account for different distributions across multiple assays, the median of the channel is used as the function to aggregate multiple matches into one.

### Join assays
Sample size is often a limiting factor in hypothesis testing. A strict quality control and the fact that TMT reagents are only available up to 18-plex can reduce the number of observed samples below the critical threshold, leading to an early end of analyses. To overcome this limitation, the provided software is capable of processing multiple runs simultaneously, allowing for testing of multiple batches and increasing the number of samples that can be included in the analysis. For further processing the runs are joined in one object at the peptide level.

### Calculate median reporter ion intensity (RI) and filter
Columns which do not meet the desired intensity can be filtered by a threshold set on the RI. The median RI can also be used to check if an entire channel has a lower detection level. This can be due two reasons. One is the expression level of the given proteins in a cell. Meaning, that the expression of the observed cell type is simply lower than the other type. Another one could be a spillage of TMT detection in other channels due incorrect or missing correction of the TMT isotopes. The correction values can be found on the vendor`s webpage (in this case Thermo Fisher) and correction needs to be set up in MaxQuant.

### Calculate median coefficient of variation (CV) and filter
Depending having a bulk sample or single-cell sample choosing a minimum of observed peptides and a cutoff value for the CV, changes the level of confidence in the peptide data. The coefficient of variation of a peptide is considered as the ratio of the standard deviation to the mean and describes the relationship of the observed peptide signal over multiple proteins (=razor proteins). Peptides having a high coefficient of variation over many razor proteins are considered as noise and need to be filtered out before statistical analysis.

@@

### Transformation of protein expression data
Data transformation applies a function to each value of a matrix or array, so that: 
$$
y_i = f(x_i)
$$
Depending on the distribution of the values in the observed expression set, different transformations can be applied to fulfill assumptions for statistical testing. In the shiny application, various procedures are implemented and can be further expanded upon request from the user. The macrophage analysis done by Specht et al. mentioned in their SCoPE2 publication \citep{Specht2021} uses the logarithm to the base 2 to spread a compacted distribution and remove skewness in the dataset. This transformation was used as a reference when comparing methods. However, using the logarithm to the base 10 may be an easy way to recalculate expression data in the mind and will also be facilitated by other transformation methods such as the boxcox method, which is also implemented in the application.
When observing a wide distribution of small and large values of the expression set in a histogram, a square root transformation can help increase the variability of smaller values and decrease the variability of larger values.
Depending on the distribution, the user has to decide which transformation to consider and verify that statistical assumptions are met after application with visualizations such as histograms, qq-plots, and MA-plots.
For statistically inexperienced users, the boxcox transformation can help with this decision. Developed in 1964 by Box and Cox \citep{Sakia1992}, this method applies a linear model against 1 on the data to determine the statistical parameter lambda by the maximum likelihood method. The log-likelihood method takes the 95% confidence interval for the parameter lambda and the final lambda is chosen as the value with the highest log-likelihood value.

@@

### Normalization
The term normalization is ambiguous in data-science and will be explained briefly in this chapter. Starting with converting the data to the Z-distribution (Gaussian result), where 0 represents the mean and 1 represents the standard deviation, in vectorization every feature (or protein) is represented as a vector that points in a specific direction in the unit sphere, normalization can have different meanings. Normalization is considered as part of the scaling procedure and ensures that every feature contributes equally to the statistical model and hinders large values from biasing the model in a particular direction. However, this could also have negative consequences for the modeling procedure, as it may reduce the impact of important features on the dataset. 
Depending on the algorithms later used in the data processing, the method of choice for normalization can change the results significantly. The K-nearest neighbor (K-NN) clustering algorithm, for instance relies on the distances between distinct data points by minimizing variance of the squared euclidean distances. This method will be applied in the missing value imputation, which will be the next step applied to the data. 
The first normalization method available in the user interface is column-wise median and row-wise mean normalization. This method was chosen by Specht et al in the SCoPE2 publication 2021 \citep{Specht2021} and worked as a reference during the development process to benchmark other methods. When dividing each value of an expression set by the median of the particular column, the procedure is considered as a column-wise median normalization. Although the mean-normalization works the same way but takes the mean as the second variable. In the analysis pipeline both procedures were applied on the dataset. However when the logarithm as the transformation method was chosen as transformation method, ProteoScanR automatically switches from a division to a subtraction, since:
$\log_a \left(\frac{u}{v}\right)=\log_a u-\log_a v$. After applying this normalization method, each value can be interpreted as a distance or fraction of the mean or median of the corresponding column or row, enabling fair comparisons between features and samples.
The second method for normalization included is the CONSTANd method, which was proven suitable for relative quantification in the field of proteomics \citep{Maes2016, VanHoutven2021}, but can also be applied on RNAseq data. CONSTANd uses a technique called matrix raking, which employs the RAS algorithm. The expression set reflects the non-negative real $(m, n)$ matrix $A$ were the bi proportional constrained matrix problem will be solved by finding the $(m, n)$ matrix $B$ which equals to $diag(x) * A *diag(y)$ . Whereas $x \in \mathbb{R}^{m}$ and $y \in \mathbb{R}^{n}$. The solution to this problem involves finding the row sum of $B=u_i$ and the column sum of $B=v_j$, where $i$ and $j$ denote the row and column indices of the matrix, respectively \citep{Bacharach1965}. In other words, the matrix will be alternatively manipulated on rows and columns until the mean of both equals 1. In order to apply the method and yield true results, one has to consider 3 major assumptions between sample types. This assumption can be observed in the MA-plot which indicates the differences of two samples. Given the intensities of two samples, $R$ and $G$, visualized on a graph with the x-axis as $M=\log_2(\frac{R}{G})$ and the y-axis as $A=\frac{1}{2}\log_2(RG)$. A comparison between all sample types of identically processed sets is needed before calculation and the assumptions must be fulfilled.

@@

### Missing value handling
When analyzing multiple batches simultaneously, it is possible to encounter situations where a particular protein is not detected in one of the batches. From a biological perspective, it is highly unlikely that a cell completely lacks a single protein, as it would imply a complete loss of function for that protein. Therefore, an intensity value of 0 is close to impossible for most proteins. In order to handle this dark space in the expression set, several methods can be chosen in the application. 
A variety of function employs replacing the missing values with the mean or median of the rest of the matrix. This procedure introduces no bias in the data if the missingness is low, although differential expression for the sample carrying the missing feature can not be expected as well. Even a more conservative approach is dropping the rows for the feature with the missing value in one of the observed samples. 
A common practice in proteomics is utilizing the K-nearest neighbor algorithm (=KNN), which imputes the missing values by a euclidean metrics of the neighboring values in the columns where the feature is not missing. $K$ denotes for the count of neighboring values of a gene and can be selected upon preference. The euclidean distance between two points is defined as $d = \sqrt{a^2 * b^2}$, the $K$ nearest neighbors average value will be assigned to the intensity of the missing feature. Since the average is a statistical metric which can be biased easily by skewed distributions, normalization is an important task before performing missing value imputation. KNN was benchmarked  for the regression of protein abundance intensities and showed an Area under the Curve (AUC) above 0.7 when testing on a human proteome test set \citep{Lan2013}.

### Batch correction 
Since mass spectrometry experiments are highly influenced by several factors, such as sample processing, the technician, manufacturer and production of the device, the reagents etc., working in multiple batches introduces further variance in the protein data. This variance would overshadow the biological differences across sample types when performing dimensionality reduction. A common misconception in biology is, that batches need to be corrected before differential expression analysis, which is the final aim of ProteoScanR. It appears convenient to remove the batch effect before performing any statistical analysis, which is in turn prone to errors especially if the sample groups are not split equally between batches. In this case, group differences influence the batch effect reducing statistical power after the correction. These both effects are so called interdependent. The opposite could happen if the experimental design is heavily unbalanced leading to group differences induced by batch correction. 
Finally it is recommended to include the batch factor should in the linear model as a co-factor and statistically quantify it \citep{Nygaard2016}.
However it could be of interest to observe biological relevance of the factor of interest in the dimensionality reduction visualization, so a batch correction option was included to meet this need.
To address the non-biological experimental variations, also known as technical factors, the R package sva \citep{Leek2012} with the ComBat function \citep{Johnson2007} was used. 
The expression of a gene $g$ for the batch $m_i$ and sample $j$ can be defined as  $Y_{ijg} = a_{g} + X\beta_{g} + \gamma_{ig} + \delta_{ig}\epsilon_{ijg}$, where $\gamma$ and $\delta\epsilon$ represent the systematic error caused by the batch. 
The Bayesian framework is employed to remove this systematic error by shrinking the effect and pooling information over the features $g$. Ideally the effects $a$ and $\beta$ remain as natural as they are and should reflect the true biological difference between experimental obtained samples or groups. Before applying batch correction, the data is standardized using Z-transformation. By the method of moments (= the mean and the variance) the two batch effect parameters $\gamma$ and $\delta\epsilon$ were estimated and can be subtracted from each value for the particular feature driving statistical moments at a comparable level between batches.
The ComBat function allows the user to specify a desired effect, which represents the actual biological variance to be preserved while removing non-biological variance. This is achieved by constructing a design matrix, which is provided as prior information about the dataset within the Bayesian framework. However, in cases of poor experimental design, the factors contributing to unwanted and desired variation may be confounded. In the worst case, if only one sample type per run is observed, it becomes challenging for the program to distinguish between batch-induced variance and biological variance. In such cases, the user is advised to adjust the experimental layout for future experiments. 
The detection algorithm for confounding effects is facilitated by QR matrix decomposition. The matrix $A=QR$ is decomposed into an orthogonal matrix $Q$ and an upper triangular matrix $R$. 
By comparing the rank of matrix $A$ with the number of biologically relevant factors, it is possible to determine the presence of confounding effects in the dataset. The rank is defined as the maximum number of linearly independent columns. The same approach is used in the limma package \citep{Phipson2016} to identify non-solvable coefficients in experimental designs.

>>>>

## Wetlab - Materials and Methods
For testing and development of the ProteoScanR pipeline the following materials and methods were applied.

### Celltypes
Macrophages from 12 healthy donors and 12 HIV patients with metabolic syndrome and 12 HIV patients without metabolic syndrome were isolated. Cell sorting was performed with FACSAria Fusion (BD Biosciences) by auto fluorescence.

@@

### Labeling
Samples of the test data set were labeled with 12-plex TMTpro (Thermo Fisher) in anhydrous acetonitril (ACN) solution.

@@

### Hardware
Data was analysed using commercially available hardware, find details below:
Lenovo Laptop X360 with:
Intel(R) Core(TM) i5-6200U CPU @ 2.30GHz
8 GB RAM

@@

### Filter by q-value
The next step for quality control is the exclusion of spectra with a high false discovery rate (FDR). When applying multiple statistical testing (e.g. t-Test) the obtained p-values can be considered as biased, because the probability to observe a significant result will iteratively increase with each test performed. Corrections in statistics are an approach to compensate for the multiplicity of testing. There are many ways to do this compensation like the Bonferroni method or Benjamini-Hochberg`s FDR. Many search engines face that problem by calculating the false discovery rate, by dividing false PSMs (=hit of the decoy database) with the total number of PSMs above the peptide-spectrum matching score. The peptides spectrum matching score is defined as $-10 * log(p)$. Whereas the p-value is defined that the hit is done by chance. The calculation of the score is highly dependent on the data acquisition method used. MaxQuant uses Andromeda \citep{Cox2011}, an integrated search engine. Proteome Discoverer from Thermo Fisher utilizes different engines such as Mascot \citep{Brosch2009}. As published by J.Cox in 2011 Mascot and Andromeda showed similar performance when comparing FDR values as a function of coverage. However the observed performance can be lower when dealing with a decreased coverage \citep{Cox2011}.The threshold for accepting an FDR of an individual PSM is described as q-value.

### Peptide spectrum match (PSM) aggregation to peptides 
In data science, aggregation refers to a row-wise operation that merges data based on a particular column using a specific function. In the context of processing of PSMs to unique peptides, the desired column is the peptide sequence. To account for different distributions across multiple assays, the median of the channel is used as the function to aggregate multiple matches into one.

### Join assays
Sample size is often a limiting factor in hypothesis testing. A strict quality control and the fact that TMT reagents are only available up to 18-plex can reduce the number of observed samples below the critical threshold, leading to diminished statistical power. To overcome this limitation, the provided software is capable of processing multiple runs simultaneously, allowing for testing of multiple batches and increasing the number of samples that can be included in the analysis. For further processing the runs are joined in one object at the peptide level.

### Calculate median reporter ion intensity (RI) and filter
Columns which do not meet the desired intensity can be filtered by a threshold set on the RI. The median RI can also be used to check if an entire channel has a lower detection level. This can be due two reasons. One is the expression level of the given proteins in a sample. Meaning, that the expression of the observed sample type is simply lower than in the other type. Another one could be a spillage of TMT detection in other channels due incorrect or missing correction of the TMT isotopes. The correction values can be found on the vendor`s webpage (in this case Thermo Fisher) and correction needs to be set up in MaxQuant.

### Calculate median coefficient of variation (CV) and filter
Depending having a bulk sample or single-cell sample choosing a minimum of observed peptides and a cutoff value for the CV, changes the level of confidence in the peptide data. The coefficient of variation of a peptide is defined as the ratio of the standard deviation to the mean and describes the relationship of the observed peptide signal over multiple proteins (=razor proteins). Peptides having a high coefficient of variation over many razor proteins are considered as noise and need to be filtered out before statistical analysis.

@@

### Transformation of protein expression data
Data transformation applies a function to each value of a matrix or array, so that: 
$$
y_i = f(x_i)
$$
Depending on the distribution of the values in the observed expression set, different transformations can be applied to fulfill assumptions for statistical testing. In the shiny application, various procedures are implemented and can be further expanded upon request from the user. The macrophage analysis done by Specht et al. mentioned in their SCoPE2 publication \citep{Specht2021} uses the logarithm to the base 2 to spread a compacted distribution and remove skewness in the dataset. This transformation was used as a reference when comparing methods. However, using the logarithm to the base 10 has the same effect and improves the interpretability of the values. Furthermore the log10 transformation will also be facilitated by other transformation methods such as the boxcox method, which is also implemented in the application.
When observing a wide distribution of small and large values of the expression set in a histogram, a square root transformation can help increase the variability of smaller values and decrease the variability of larger values.
Depending on the distribution, the user has to decide which transformation to consider and verify that statistical assumptions are met after application with visualizations such as histograms, qq-plots, and MA-plots.
For statistically inexperienced users, the boxcox transformation can help with this decision. Developed in 1964 by Box and Cox \citep{Sakia1992}, this method applies a linear model against 1 on the data to determine the statistical parameter lambda by the maximum likelihood method. The log-likelihood method takes the 95% confidence interval for the parameter lambda and the final lambda is chosen as the value with the highest log-likelihood value.

@@

### Normalization
The term normalization is ambiguous in data-science and will be explained briefly in this chapter. Starting with converting the data to the Z-distribution (Gaussian result), where 0 represents the mean and 1 represents the standard deviation, in vectorization every feature (or protein) is represented as a vector that points in a specific direction in the unit sphere, normalization can have different meanings. Normalization is considered as part of the scaling procedure and ensures that every feature contributes equally to the statistical model and hinders large values from biasing the model in a particular direction. However, this could also have negative consequences for the modeling procedure, as it may reduce the impact of important features on the dataset. 
Depending on the algorithms later used in the data processing, the method of choice for normalization can change the results significantly. The K-nearest neighbor (K-NN) clustering algorithm, for instance relies on the distances between distinct data points by minimizing variance of the squared euclidean distances. This method will be applied in the missing value imputation, which will be the next step applied to the data. 
The first normalization method available in the user interface is column-wise median and row-wise mean normalization. This method was chosen by Specht et al in the SCoPE2 publication 2021 \citep{Specht2021} and worked as a reference during the development process to benchmark other methods. When dividing each value of an expression set by the median of the particular column, the procedure is considered as a column-wise median normalization. Although the mean-normalization works the same way but takes the mean as the second variable. In the analysis pipeline both procedures were applied on the dataset. However when the logarithm as the transformation method was chosen as transformation method, ProteoScanR automatically switches from a division to a subtraction, since:
$\log_a \left(\frac{u}{v}\right)=\log_a u-\log_a v$. After applying this normalization method, each value can be interpreted as a distance or fraction of the mean or median of the corresponding column or row, enabling fair comparisons between features and samples.
The second method for normalization included is the CONSTANd method, which was proven suitable for relative quantification in the field of proteomics \citep{Maes2016, VanHoutven2021}, but can also be applied on RNAseq data. CONSTANd uses a technique called matrix raking, which employs the RAS ($K=RAS$) algorithm. The expression set reflects the non-negative real $(m, n)$ matrix $A$ were the bi proportional constrained matrix problem will be solved by finding the $(m, n)$ matrix $K$ which equals to $diag(x) * A *diag(y)$, so that $K=diag(x) * A *diag(y)= RAS$. Whereas $x \in \mathbb{R}^{m}$, $y \in \mathbb{R}^{n}$ and $R$ denotes for the row multiplier and $S$ for the column multiplier. The solution to this problem involves finding the row sum of $K=u_i$ and the column sum of $K=v_j$, where $i$ and $j$ denote the row and column indices of the matrix, respectively \citep{Bacharach1965}. In other words, the matrix will be alternatingly manipulated on rows and columns until the mean of both equals 1. In order to apply the method and yield true results, one has to consider 3 major assumptions between sample types. These assumption can be observed in the MA-plot which indicates the differences of two samples. Given the intensities of two samples, $R$ and $G$, visualized on a graph with the x-axis as $M=\log_2(\frac{R}{G})$ and the y-axis as $A=\frac{1}{2}\log_2(RG)$. A comparison between all sample types of identically processed sets is needed before calculation and the assumptions must be fulfilled.

@@

### Missing value handling
When analyzing multiple batches simultaneously, it is possible to encounter situations where a particular protein is not detected in one of the batches. In order to handle this dark space in the expression set, several methods can be chosen in the application. 
A variety of function employs replacing the missing values with the mean or median of the rest of the matrix (total mean/median). This procedure introduces no bias in the data if the missingness is low, although differential expression for the sample carrying the missing feature can not be expected as well. An even more conservative approach is dropping the rows for the feature with the missing value in one of the observed samples. 
A common practice in proteomics is utilizing the K-nearest neighbor algorithm (=KNN), which imputes the missing values according to an euclidean metric of the neighboring values in the columns where the feature is not missing. $K$ denotes for the count of neighboring values of a protein which are considered and can be selected upon preference. The euclidean distance between two points is defined as $d = \sqrt{a^2 * b^2}$, the $K$ nearest neighbors average value will be assigned to the intensity of the missing feature. Since the average is a statistical metric which can be biased easily by skewed distributions, normalization is an important task before performing missing value imputation. KNN was benchmarked  for the regression of protein abundance intensities and showed an Area under the Curve (AUC) above 0.7 when testing on a human proteome test set \citep{Lan2013}.

### Batch correction 
Since mass spectrometry experiments are highly influenced by several factors, such as sample processing, the technician, manufacturer and production of the device, the reagents etc., working in multiple batches introduces further variance in the protein data. This variance can overshadow the biological differences across sample types when performing dimensionality reduction. A common misconception in biology is, that batches need to be corrected before differential expression analysis, which is the final aim of ProteoScanR. It appears convenient to remove the batch effect before performing any statistical analysis, which is in turn prone to errors especially if the sample groups are not split equally between batches. In this case, group differences influence the batch effect reducing statistical power after the correction. These both effects are so called interdependent. The opposite could happen if the experimental design is heavily unbalanced leading to group differences induced by batch correction. 
Finally it is recommended to include the batch factor should in the linear model as a co-factor and statistically quantify it \citep{Nygaard2016}.
However it could be of interest to observe biological relevance of the factor of interest in the dimensionality reduction visualization, so a batch correction option was included to meet this need.
To address the non-biological experimental variations, also known as technical factors, the R package sva \citep{Leek2012} with the ComBat function \citep{Johnson2007} was used. 
The expression of a gene $g$ for the batch $m_i$ and sample $j$ can be defined as  $Y_{ijg} = a_{g} + X\beta_{g} + \gamma_{ig} + \delta_{ig}\epsilon_{ijg}$, where $\gamma$ and $\delta\epsilon$ represent the systematic error caused by the batch. 
The Bayesian framework is employed to remove this systematic error by shrinking the effect and pooling information over the features $g$. Ideally the effects $a$ and $\beta$ remain as natural as they are and should reflect the true biological difference between experimental obtained samples or groups. Before applying batch correction, the data is standardized using Z-transformation. By the method of moments (= the mean and the variance) the two batch effect parameters $\gamma$ and $\delta\epsilon$ were estimated and can be subtracted from each value for the particular feature driving statistical moments at a comparable level between batches.
The ComBat function allows the user to specify a desired effect, which represents the actual biological variance to be preserved while removing non-biological variance. This is achieved by constructing a design matrix, which is provided as prior information about the dataset within the Bayesian framework. However, in cases of poor experimental design, the factors contributing to unwanted and desired variation may be confounded. In the worst case, if only one sample type per run is observed, it becomes challenging for the program to distinguish between batch-induced variance and biological variance. In such cases, the user is advised to adjust the experimental layout for future experiments. 
The detection algorithm for confounding effects is facilitated by QR matrix decomposition. The matrix $A=QR$ is decomposed into an orthogonal matrix $Q$ and an upper triangular matrix $R$. 
By comparing the rank of matrix $A$ with the number of biologically relevant factors, it is possible to determine the presence of confounding effects in the dataset. The rank is defined as the maximum number of linearly independent columns. The same approach is used in the limma package \citep{Phipson2016} to identify non-solvable coefficients in experimental designs.

### 2023-07-18 markdown/results.Rmd
until batch correction done

After a runtime of 2 minutes on the 4 core laptop the data is ready to visualize. On the left side of the user interface (see figure \@ref(fig:ui-summary)), users can find the analysis pipeline settings that can be adjusted. Once the desired changes are made, they can be applied by clicking the update button. This allows users to observe the direct impact of the applied changes on the dataset.
In the summary visualization, it may appear cluttered due to the processing of 36 files. In smaller datasets, the nodes of the network graph can be more easily distinguished and understood.
When running the application with SCoPE2 \citep{Specht2021} reference cutoffs, the following results were visualized.

\newpage
```{r ui-numbers-default, echo=F, out.width="100%", fig.align='center', fig.cap="Barchart showing the number of peptide spectrum matches, peptides and proteins"}
knitr::include_graphics("screenshots/numbers_default.png")
```
In order to get an idea of the magnitude of the data set, the barplot provides an overview (see figure \@ref(fig:ui-numbers-default))
The default settings of the analysis pipeline resulted in a total number of unique proteins below approximately 1,000, which were aggregated from around 3,000 peptides. Before these peptides were aggregated over approximately 32,000 peptide spectrum matches. The aggregation for both steps was performed by the column median.

@@

\newpage
```{r ui-constand-norm, echo=F, out.width="100%", fig.align='center', fig.cap="Popup showing the dependencies for the CONSTANd normalization"}
knitr::include_graphics("screenshots/constand_norm_dependency.png")
```

@@

\newpage
```{r ui-RI-st-default, echo=F, out.width="100%", fig.align='center', fig.cap="Reporter ion intensity with selection on sampletype"}
knitr::include_graphics("screenshots/RI_st_default.png")
```
Comparing sample types in regards of reporter ion intensity is recommended to observe any experimental bias such as batch/channel confounding effects which would influence the statistical model afterwards.

@@

\newpage
```{r ui-cv-st-default, echo=F, out.width="100%", fig.align='center', fig.cap="Median covariance of razor proteins with selection on sampletype"}
knitr::include_graphics("screenshots/cv_st_default.png")
```
The distribution of median coefficient of variation (CV) shows the uniqueness in peptides for their respective razor peptides. This metric represents the presence or absence of noisy matches in the data set.

@@

\newpage
```{r ui-cv-rawfile-default, echo=F, out.width="100%", fig.align='center', fig.cap="Median coefficient of variation (CV) of leading razor proteins per cell with selection on Raw.file"}
knitr::include_graphics("screenshots/cv_rawfile_default.png")
```
Razor peptide ambiguity over multiple batches can indicate experimental failures during sample prep or during acquisition.

>>>>

After a runtime of 2 minutes on the 4 core laptop the data is ready to visualize. On the left side of the user interface (see figure \@ref(fig:ui-summary)), users can find the analysis pipeline settings that can be adjusted. Once the desired changes are made, they can be applied by clicking the update button. This allows users to observe the direct impact of the applied changes on the dataset.
In the summary visualization, it may appear cluttered due to the processing of 36 files. In smaller data sets, the nodes of the network graph can be more easily distinguished and understood.
When running the application with SCoPE2 \citep{Specht2021} reference cutoffs, the following results were visualized.

\newpage
```{r ui-numbers-default, echo=F, out.width="100%", fig.align='center', fig.cap="Barchart showing the number of peptide spectrum matches, peptides and proteins"}
knitr::include_graphics("screenshots/numbers_default.png")
```
In order to get an idea of the magnitude of the data set, the barplot provides an overview (see figure \@ref(fig:ui-numbers-default)).
The default settings of the analysis pipeline resulted in a total number of unique proteins below approximately 1,000, which were aggregated from around 3,000 peptides. Before these peptides were aggregated over approximately 32,000 peptide spectrum matches. The aggregation for both steps was performed by the column median.

@@

\newpage
```{r ui-constand-norm, echo=F, out.width="100%", fig.align='center', fig.cap="Popup showing the dependencies for the CONSTANd normalization"}
knitr::include_graphics("screenshots/constand_norm_dependency.png")
```
Choosing the CONSTANd option for normalization a popup window appears to test for assumptions on the dataset.

@@

\newpage
```{r ui-RI-st-default, echo=F, out.width="100%", fig.align='center', fig.cap="Reporter ion intensity with selection on sampletype"}
knitr::include_graphics("screenshots/RI_st_default.png")
```
Comparing sample types in regards of reporter ion intensity is recommended to observe any experimental bias such as batch/channel confounding effects which would negatively impact the statistical model afterwards.

@@

\newpage
```{r ui-cv-st-default, echo=F, out.width="100%", fig.align='center', fig.cap="Median covariance of razor proteins with selection on sampletype"}
knitr::include_graphics("screenshots/cv_st_default.png")
```
The distribution of median coefficient of variation (CV) shows the uniqueness in peptides for their respective razor protein. This metric represents the presence or absence of noisy matches in the data set.

@@

\newpage
```{r ui-cv-rawfile-default, echo=F, out.width="100%", fig.align='center', fig.cap="Median coefficient of variation (CV) of leading razor proteins per cell with selection on Raw.file"}
knitr::include_graphics("screenshots/cv_rawfile_default.png")
```
Razor peptide ambiguity over multiple batches can indicate experimental failures during sample prep (especially digestion) or during acquisition.
